---
title: "p8157_hw1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(Hotelling)
```

# P8157 - Analysis of Longitudinal Data Homework - 1

## Part A

1. Plasma inorganic phosphate measurements obtained from 13 control and 20 obese patients 0, 0.5, 1, 1.5, 2, 
and 3 hours after an oral glucose challenge. The investigators intend to test the following hypotheses using 
Hotelling’s T2 statistic:

(a) To test the null hypothesis that the group means are the same at all six measurement times. 

$H_0: $


(b) To test whether the profiles in the two groups are parallel.

$H_0: \textbf{ABC=D}$. Profiles in the two groups are parallel i.e. no group-time interaction, the slopes between 
each time points are identical between the two groups:

$$
\mu_{12}-\mu_{11} = \mu_{22} - \mu_{21} \\
\mu_{13}-\mu_{12} = \mu_{23} - \mu_{22} \\
...\\
\mu_{16} - \mu_{15} = \mu_{26} - \mu_{25}
$$

$$
\begin{equation*}
B_{2, 6} = 
\begin{bmatrix}
\mu_{11} & \mu_{12} & ... & \mu_{16} \\
\mu_{21} & \mu_{22} & ... & \mu_{26} 
\end{bmatrix}
\end{equation*}
$$
$$
\begin{equation*}
C_{6, 5} = 
\begin{bmatrix}
-1 & 0 & ... & 0 \\
1 & -1 & ... & 0 \\
0 & 1 & ... & 0 \\
... \\
0 & 0 & ... & 1
\end{bmatrix}
\end{equation*}
\\
$$

(c) To test whether the differences in means at 2 and 3 hours after an oral glucose challenge are different between the control and obese patients.

```{r 1_c}

```

## Part B



```{r B_1}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
